---
title: "Experimental Design for HUD FAFSA 3"
author: "Michael DiDomenico and Nathaniel Higgins"
output: html_document
---

# Purpose
The purpose of this document is to explain (and implement) the randomization strategy
for the HUD-FAFSA-3 project.

# Necessary Background
HUD needs to develop evidence to support an existing grant mechanism which distributes
money to Public Housing Authorities (PHAs) to hire "navigators" -- individuals who
assist HUD beneficiaries living in PHAs with their transition from high school to 
college. If there is promising evidence, the existing grant mechanism is more likely 
to get continuing appropriations. The more immediate outcomes to residents could be 
better educational outcomes (more FAFSAs completed, more aid, higher post-secondary 
attendance, better institutional fit). We will evaluate using the best available design.

In the current granting period, a total of 9 PHAs received funding:

1. Prichard, AL
2. Pheonix, AZ
3. Los Angeles, CA
4. Northwest Georgia
5. Chicago, IL
6. High Point, NC
7. Philadelphia, PA
8. Seattle, WA
9. Milwuakee, WI

Four of these PHAs are too small to include in a randomized evaluation and will be 
excluded from the remainder of the random assignment discussion (Prichard, NW Georgia, 
High Point, and Phoenix). In each of the four, navigators will attempt to provide 
assistance to all residents in the targeted age range. HUD and OES are exploring the 
possibilities of non-experimental comparisons for these PHAs.

## Grant requirements 
Each PHA has received grant funding to hire at least one navigator. The navigator(s) will 
attempt to serve residents 15 - 20 years old) living in public housing (including 
scattered-site housing within the PHA. The population of interest lives in PHA-managed 
buildings that are managed at an administrative unit called the "AMP". The grant 
specifies that PHAs will agree to serve AMPs randomly assigned by HUD to receive treatment


An AMP usually contains multiple buildings. Buildings may be geographically clustered 
or spread out across scattered-site housing. Furthermore, in some cases multiple AMPs 
could comprise one development with several phases of construction. 

Given the diversity of AMP-types, OES relied heavily on local PHA knowledge to form 
randomization clusters. Each PHA provided HUD with a list of certain AMPs they requested 
be removed from consideration for various reasons, for example due to geography, overlap
with other resident programs (e.g., Jobs Plus), or other local knowledge. In a majority 
of cases, individual AMPs are given their own randomization clusters. However, we have 
grouped certain AMPs where PHAs expressed a preference for doing so. 

The final lists of AMPs and clusters are provided by PHAs and may not match the most recent
information included in the Public and Indian Housing Information Center (PIC) database. 
For example, several PHAs have re-organized AMPs creating new AMPs and discontinuing the 
use of older AMP ids. Therefore, it will be vital to make sure there is a mechanism in 
place for creating individual-level datasets that correspond to the randomization clusters 
for matching to outcome data (e.g., records on FAFSA completion held by the Department of 
Education).

# Experimental Assignment
Our experimental design will distribute AMPs within PHAs to treatment. That is, we
will designate treatment AMPs, and instruct navigators to treat as many
age-eligible youth within the treatment AMP as they are able to. We will observe outcomes
at the AMP-level (for example, we will observe the proportion of age-eligible youth
within each AMP who complete the FAFSA). Each AMP included in the study will be either
a treatment AMP (navigators instructed to treat as many individuals as possible within
that AMP) or a control AMP (navigators instructed *not* to treat any individuals living
within that AMP).

In each PHA where there is more than one navigator, we will not actually dictate which
navigator will be assigned to which treatment AMP. Instead, we will allow PHAs to
hire and distribute navigators to treatment AMPs as they see fit. We will, however,
attempt to prevent any distribution of treatment AMPs that would be infeasible for
the given number of navigators to serve (see below under "Important Considerations").

## Important Considerations
We could, in principle, choose AMPs using simple randomization. There are several
complications that restrict us from using simple randomization on the AMPs listed
above. We list them here:

1. Because AMPs are of varying size, and because we are limited in the number of
navigators within PHA, the probability of treatment will not be uniform between PHAs.
The probability of treatment will be identical within blocks of AMPs, but different
for AMPs included in the study that are from different PHAs. Implication: we will
block by PHA and will include PHA fixed effects in the model, using only within-PHA 
variation to identify treatment effects.

2. Because AMPs are of varying size, and because we will only be able to observe
important outcomes at the AMP level, we will seek to "flood" AMPs selected for treatment,
such that a treatment AMP has sufficient staffing by navigators that the navigators
can serve as many individuals as desire service. Implication: we will (1) exclude from
the study certain AMPs that are too large and (2) rather than randomizing a set number
of AMPs to treatment and control, we will continue to assign AMPs to treatment until
each navigator is fully engaged ("busy"). We will use randomization inference to 
evaluate all possible assignments to treatment under this unusual distribution method
(any other ideas?).


```{r echo=FALSE}
# Basic parameters (these are easily changed)
# Since this program uses random numbers, we'll want to start with setting the seed 
# so that our work is replicable
set.seed(3472) # 3472 is random integer chosen using random.org:
# Timestamp: 2017-03-31 17:16:42 UTC

# Overall individual takeup rate ("rate") for offered services:
rate <- 0.5 # Set at 0.5 as a baseline

# want to make this adjustable for each PHA
# List of PHAs and their number of navigators and max load for each. I am giving slightly 
# smaller caseloads to HACLA and SHA because of there relatively fewer clusters and 
# number of youth, respectively
state <- c("CA", "IL", "PA", "WA", "WI")
nav <- c(3, 3, 2, 3, 1)
# Max workload ("ml") per navigator
ml  <- c(125,150,150,82,125)

pha.params <- cbind.data.frame(state,nav,ml)

# call in random assignment library with function that will be used
library(hudfafsaiii)
library(knitr)
```

The number of navigators to be funded by each PHA, the number of AMPs in which youth
served by the navigators are distributed, the number of those residents to be served 
(estimated using data provided by PHAs in most cases, but by PIC for Los Angeles), the 
number of residents who would need to be contacted by the navigators (assuming a takeup 
rate of the navigator's services of `r rate`), and the total number of residents in the 
relevant age range are displayed in the table below. Finally, we also display our estimate of 
the percentage of total residents that the navigators would need to contact in order 
to be fully engaged (i.e. "busy") assuming a maximum workload of `r ml` students.

# Randomization Procedure
Taking into account the considerations above, we propose the following randomization
procedure.

- Stage 1:
For each PHA, we will eliminate from consideration AMPs that are too large or too small.
The smallest AMPs -- those with fewer than 10 age-eligible individuals -- are not
worth traveling to and will be excluded from the study. We will also exclude those 
AMPs that are so large that the number of age-eligible youth expected to engage navigator 
services would be more than a single navigator could handle. 

- Stage 2a:
We will sort AMPs by a random number, assigning the first AMP to treatment, then

- Stage 2b:
Assigning subsequent AMPs to treatment until the current AMP assigned to treatment 
exceeds the maximum workload, then

- Stage 3:
Assigning all remaining AMPs to control.


## Threats to the strategy
XXX


# PHAs with the number of navigagors, clusters, and residents
```{r}
# Load data
clusters.with.amps <- read.csv(paste(mywd,"clusters.with.amps.csv",sep=""))

# aggregate total by OES cluster
clusters <- aggregate(tot_age_eligible ~ OES_cluster + state, data=clusters.with.amps, FUN=sum)

# merge to navigator/workload parameters
dta <- merge(clusters, pha.params, by="state", all.x=T)

# Create new variables needed for randomization
# Calculate total number of age-eligible to-be-served in each AMP
dta$tot.served <- dta$tot_age_eligible*rate
# Total workload of navigator(s):
dta$tot.load <- 0
# Create a treatment indicator
dta$treat <- 0

# Remove all clusterss w/ fewer than 10 age-eligible youth
dta <- dta[dta$tot_age_eligible >= 10, ]

# Remove all AMPs w/ too many age-eligible youth to be serviced by a single navigator
# Not applying to HACLA after consulting with the PHA and HUD
dta <- subset(dta, state == "CA" | dta$tot_age_eligible <= as.numeric(dta$ml)/rate)

# Number of AMPs in population
(N <- dim(dta)[1])


# this table should be updated to reflect taking out the AMPs PHAs want removed
table1 <- as.data.frame(as.list(aggregate(tot_age_eligible ~ state + nav + ml, data=dta, FUN=function(x) c(length(x),sum(x)))))
table1$to.serve <- table1$nav*table1$ml
table1$to.contact <- table1$to.serve/rate
table1$pct.to.contact <- table1$to.contact/table1$tot_age_eligible.2*100
table1$pct.to.contact <- sprintf("%3.1f%%", table1$pct.to.contact)
names(table1) <- c("State","Navigators","Maximum per Navigator","Clusters","Population","To Serve","To Contact","Percent to Contact")
kable(table1[order(table1$State),],row.names=F)
```

Although almost all residents could be contacted in Seattle given the number of navigators,
unlike Phoenix which had only 5 clusters, Seattle has 15 clusters and a fair number are scattered
site. We will assign each navigator fewer students in Seattle to attempt to account for this.


### Assign a random number to each AMP in the population and sort
```{r}
# Random number
dta$rand <- runif(n = N)

# Sort by random number within PHA
dta <- dta[order(dta$state, dta$rand), ]
```


### "Regular" PHAs: Chicago, Philly, and Los Angeles
We will cycle through Chicago, Philly, and Los Angeles using the same assignment 
strategy. We will address Phoenix, Seattle, and Wisconsin separately (Phoenix and
Seattle because of their small number of AMPs and small population; Wisconsin because 
it only has one navigator so travel time between locations becomes more of an issue 
without the possibility of dividing up geography).

```{r}
# call the function we created to make the assignments
substates <- assignment(dta, pha.params = pha.params)

# save the assignments (this is the minimum info necessary, but we can add to it. we
# may also want to merge back to the original spreadsheets to give these back to phas
# at the amp level)
substates.amps <- merge(clusters.with.amps,substates[,c("OES_cluster","treat")],
   by="OES_cluster",all.X=T)

# This exports with amps so it is easier for PHAs to know where the navigators will serve
# Seattle is different because Seattle amps break across clusters, so they will need to 
# use the spreadsheet with the high school districts, so they know how to break apart
# scattered site amps.

write.csv(substates.amps,paste(mywd,"hud.fafsa.phase3.assignments.amps.csv",sep=""))

```
